Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-09-26 paper/current/main.tex
Add the new airport figure into the paper, adjust extant results text accordingly

The embedding distance correlates with and can be used to predict real-wold mobility flows better than geographic distance. 
The amount of variance in the expected flux explained by embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds when considering the subset of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
We fit the gravity model given by Eq.~\ref{eq:gravity_basic} (see Data and Methods) and computed the expected flux  $\hat{T}_{ij}$ using geographic distance and embedding distance as parameters.
The embedding distance produces better predictions of actual flux than does geographic distance (Fig.~\ref{fig:gravity_performance}c--d), which also holds for the subsets of only domestic (Fig.~\ref{fig:supp:gravity_correlation}e--f) and only international mobility (Fig.~\ref{fig:supp:gravity_correlation}g--h).
The embedding distance defined from the cosine distance between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion-based'' network distance measures such as the personalized-page rank scores (see \SI) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Our results unequivocally demonstrate that the embedding distance better captures actual mobility patterns than does geographic distance.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts mobility.}
		\textbf{a-b.} Embedding distance (cosine distance between organization vectors) better explains the expected flux between organization pairs than geographic distance.
		The red line is the line of the best fit.
		Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted, but are too small to be visible.
		Correlation is calculated on the data in the log-log scale. 
		The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{c-d.} Predictions of flux between organization pairs made using embedding distance outperform those made using geographic distance.
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$;
		a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
		\textbf{e-f.} Predictions made using model parameters learned from a subset of domestic (within-country) organization pairs.
		Embedding distance outperforms those made for geographic distance.
		\textbf{g-h.} Embedding distance also outperforms geographic distance when predicting flux using model parameters estimated from only international mobility (across-country organization pairs).
		Predictions of domestic and international mobility are qualitatively similar whether predicted using models with only within- or across-country pairs (\textbf{e-h}), or all organizations pairs (Fig. \ref{fig:supp:gravity_prediction_subsets}).
	}
	\label{fig:gravity_performance}
\end{figure}

@@

\subsection*{Data -US Airport Itinerary}
We use US airport itinerary data from the Bureau of Transportation statistics provided by the United States Department of Transportation. Origin and Destination Survey (DB1B) is a 10\% sample of domestic airline tickets, including origin, destination, and other itinerary details. Trajectories are represented as the list of airport code and cover itinerary between 1993 to 2020. There are 30,776,0841 trajectories in total and the unique number of airports is 838.

>>>>

The embedding distance correlates with and can be used to predict real-wold mobility flows better than geographic distance. 
The amount of variance in the expected flux explained by embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds when considering the subset of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
We fit the gravity model given by Eq.~\ref{eq:gravity_basic} (see Data and Methods) and computed the expected flux  $\hat{T}_{ij}$ using geographic distance and embedding distance as parameters.
The embedding distance produces better predictions of actual flux than does geographic distance (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}), which also holds for the subsets of only domestic and only international mobility.
The embedding distance defined from the cosine distance between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion-based'' network distance measures such as the personalized-page rank scores (see \SI) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Our results unequivocally demonstrate that the embedding distance better captures patterns of actual scientific mobility than does geographic distance.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts global scientific mobility and U.S flight itineraries.}
		\textbf{a-b.} Embedding distance (cosine distance between organization vectors) better explains the expected flux of researchers between organization pairs than geographic distance.
		The red line is the line of the best fit.
		Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted, but are too small to be visible.
		Correlation is calculated on the data in the log-log scale. 
		The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{c-d.} Predictions of flux between organization pairs made using embedding distance outperform those made using geographic distance.
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$;
		a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
		\textbf{e-f.} Embedding distance better explains the expected flux of passengers between U.S. airports than does geographic distance. 
		\textbf{g-h.} Predictions of flux between airport pairs made using embedding distance outperform those made using geographic distance. 
	}
	\label{fig:gravity_performance}
\end{figure}

@@

\subsection*{Data -US Airport Itinerary}
We use US airport itinerary data from the Bureau of Transportation statistics provided by the United States Department of Transportation. Origin and Destination Survey (DB1B) is a 10\% sample of domestic airline tickets, including origin, destination, and other itinerary details. Trajectories are represented as the list of airport code and cover itinerary between 1993 to 2020. There are 307,760,841 trajectories in total and the unique number of airports is 838.
